### 2022-06-20 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betacov_mapping@33a1b0ced81c54a6379e039b524e45bfb8aff78d 🚀

Disease emergence is complex, and is driven not only by animal-human
contact, but also by the underlying evolutionary dynamics in viral
reservoirs (Plowright et al. 2017). Although host richness is often used
as a superficial proxy for spillover risk (Anthony et al. 2017,
Ruiz-Aravena et al. 2022), these approaches oversimplify the relevant
interspecific heterogeneity in immunology, behavior, and other traits,
and therefore overlook unique host pools that allow for the rapid
evolution of highly divergent viruses (Agosta et al. 2010). In the case
of generalist pathogens like betacoronaviruses, there is conceptual and
empirical support to the idea that these community-level mechanisms are
even more important (Power and Mitchell 2004). This creates a disconnect
between coevolutionary theory and existing ecological frameworks for
mapping spillover risk.

@@

Here, we apply the GMTC to explore and explain the global biogeography
of betacoronaviruses, the group that includes SARS-CoV, MERS-CoV, and
SARS-CoV-2. Working from a recently published database of bat hosts of
betacoronaviruses, we develop the first global maps of both host and
virus evolutionary distinctiveness and biogeographic regions for this
system. Aiming to explain these patterns, we develop a generalized
framework for applying the GMTC to host-virus interactions, with a
specific emphasis on the potential to create independent coevolutionary
dynamics (and therefore spatial fragmentation in risk) through
heterogeneity. We develop a trivariate risk assessment system that
connects each GMTC mechanism to a quantifiable aspect of host-virus
interactions: (i) viral sharing rates in host communities, representing
the strength of potential interaction between viruses and any one host
(i.e., places where viruses undergo constant host switching may be
coevolutionary coldspots); (ii) the phylogenetic diversity of hosts, as
a proxy for variation in the immunological mechanisms that antagonize
viruses (i.e., the selection mosaic); and (iii) the local uniqueness of
the bat community, representing the potential for viruses to be exposed
to novel host traits (e.g., variation in receptor sequences). Together,
we argue that these can be used to identify and map the evolutionary
drivers that---in conjunction with transmission processes (e.g., viral
prevalence in reservoirs and animal-human contact rates)--- determine
disease emergence risk.

>>>>

Disease emergence is complex, and is driven not only by animal-human
contact, but also by the underlying evolutionary dynamics in viral
reservoirs (Plowright et al. 2017). Although host richness is often used
as a superficial proxy for spillover risk (Anthony et al. 2017,
Ruiz-Aravena et al. 2022), these approaches oversimplify the relevant
interspecific heterogeneity in immunology, behavior, and other traits,
and therefore overlook unique host pools that allow for the rapid
evolution of highly divergent viruses (Agosta et al. 2010). In the case
of generalist pathogens like betacoronaviruses, there is conceptual and
empirical support to the idea that these community-level mechanisms are
even more important (Power and Mitchell 2004), particularly given that
cross-species transmission may, as a rule, structure viral evolution
more than co-divergence with hosts {[}Geoghegan{]}. This creates a
disconnect between coevolutionary theory (including empirical evidence
from virology) and most existing ecological frameworks for mapping
spillover risk.

@@

Here, we apply the GMTC to explore and explain the global biogeography
of betacoronaviruses, the group that includes SARS-CoV, MERS-CoV, and
SARS-CoV-2. In their bat reservoirs, coronaviruses evolve through a mix
of host jumps, recombination among disparate lineages, and, to a lesser
degree, co-divergence with their hosts {[}Anthony{]}---a mix of
mechanisms that creates a complex and nonlinear relationship between
host diversity and viral emergence. Working from a recently published
database of bat hosts of betacoronaviruses, we develop the first global
maps of both host and virus evolutionary distinctiveness and
biogeographic regions for this system. Aiming to explain these patterns,
we develop a generalized framework for applying the GMTC to host-virus
interactions, with a specific emphasis on the potential to create
independent coevolutionary dynamics (and therefore spatial fragmentation
in risk) through heterogeneity. We develop a trivariate risk assessment
system that connects each GMTC mechanism to a quantifiable aspect of
host-virus interactions: (i) viral sharing rates in host communities,
representing the strength of potential interaction between viruses and
any one host (i.e., places where viruses undergo constant host switching
may be coevolutionary coldspots); (ii) the phylogenetic diversity of
hosts, as a proxy for variation in the immunological mechanisms that
antagonize viruses (i.e., the selection mosaic); and (iii) the local
uniqueness of the bat community, representing the potential for viruses
to be exposed to novel host traits (e.g., variation in receptor
sequences). Together, we argue that these can be used to identify and
map the evolutionary drivers that---in conjunction with transmission
processes (e.g., viral prevalence in reservoirs and animal-human contact
rates)--- determine disease emergence risk.
